Further Evidence of Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2022-03-28 src/ms.tex
adding weights figure

To further explore whether our \rocrit inference procedure is biased by systematics in the data, we performed non-linear least-squares fits of two models to the CKS long-period pile-up and the \lamostmcq 90th percentile curve (both in the \teff range of 5800--6250~K) using the \texttt{curve\_fit} function in \texttt{scipy.optimize}. The first model assumes \rocrit = 2 and has one free parameter, a temperature shift added to the data. The second model assumes no temperature shift and allows \rocrit to vary (\rocrit being the only free parameter). The results of these fits are shown in Figure~\ref{fig:rocrit}. We found that for the CKS sample, particularly when using the more precise temperatures from \citet{Fulton2018}, that the $\Delta \chi^2$ between the \rocrit = 2 and variable \rocrit models was negligible. In other words, the \rocrit = 2 model fits the data well with a \teff shift of $\approx$~80--150~K depending on the \teff source used. For the \lamostmcq sample, there is weak support for the variable \rocrit model. However, as shown in Appendix~\ref{app:teff}, there is a strong systematic trend in the LAMOST temperatures when compared to other spectroscopic catalogs (particularly SPOCS). Correcting for these systematic trends weakens the support for the variable \rocrit model, and we speculate that temperature systematics bias the \rocrit inference procedure. Lending further support to this hypothesis is the observation of visibly different slopes for the long-period pile-up in the CKS vs. \lamostmcq samples (see Fig.~\ref{fig:percentiles}).\footnote{\edit1{While there is reasonably good agreement between the CKS long-period pile-up and the \lamostmcq 90th percentile curve, we observe in Figure~\ref{fig:percentiles} a systematic trend in which most CKS pile-up stars lie above the \lamostmcq curve for \teff~$\lesssim$~6000~K, and below it for \teff~$\gtrsim$~6000~K. This is consistent with there being a \teff-dependent systematic trend between the \teff scales of CKS and LAMOST, as discussed in Appendix~\ref{app:teff}.}}

@@

\subsection{Detection biases}
\label{subsec:detectionbias}
\edit1{To this point, we have not addressed the detection biases inherent to Kepler rotation period catalogs. As rotation periods become longer, it becomes more difficult to measure periods directly from rotational brightness modulations in Kepler time series. This is true for at least three reasons: (1) the Kepler baseline is finite, (2) photometric variation amplitude declines with increasing rotation period, and (3) photometric variations tend to become less periodic with increasing rotation period. This is important because, if a period detection threshold is Ro-dependent and the detection threshold Ro$_\mathrm{thresh}$ is close to the value \rocrit at which WMB becomes important, then the location of the pile-up in the catalogs considered here may depend sensitively on how detectability depends on Ro. For example, if Ro$_\mathrm{thresh}$ is only slightly larger than \rocrit and the long-period pile-up has some non-negligible width in period space (due either to astrophysics or measurement error), then it's possible the current data set only reveals the lower edge of the long-period pile-up.}

@@

@@

As discussed in \S\ref{subsec:rossby} and \S\ref{subsec:models}, the existence of a pile-up of stars with a constant Rossby number is consistent with expectations from the WMB model of \citet{vanSaders2016, vanSaders2019}. The location of the long-period pile-up coincides closely with the cluster of main-sequence stars with asteroseismic rotation rates determined by \citet{Hall2021} and we suggest the two features are one and the same. Offsets in both observational and theoretical \teff scales between different studies can explain the modest discrepancies between the asteroseismic sample and the samples considered here, as well as discrepancies in the \rocrit values inferred here and those proposed in \citet{vanSaders2019}.

Furthermore, we note that there are important differences between the samples considered here and the \citet{Hall2021} asteroseismic sample. The present samples have rotation periods detected from photometric variations and, for a survey of finite baseline and sensitivity like Kepler, photometric rotation periods are harder to detect for slower rotators, smaller amplitude variations, and more stochastic variability patterns. Exacerbating the problem of detection, slower rotation tends to be accompanied by smaller-scale and less periodic variability. The asteroseismic sample, by comparison, is not biased against quiet, unspotted stars and is more likely to contain pile-up stars that our sample may have missed.

It is evident from Figure~\ref{fig:gap} that the long-period pile-up also coincides with a steep gradient in variability amplitude, such that a hidden population of pile-up stars may lie just beyond the edge of detectability. This effect, in addition to the \teff offsets mentioned above, can further bias our inference of \rocrit to lower values. Regardless of the biases mentioned above and the true value of \rocrit, the sample considered clearly indicates that the long-period pile-up is apparent for stars with \teff~$\gtrsim$~\teffmin with Rossby numbers lower than those associated with a detection edge. \edit1{In other words, the fact that we observe a pile-up indicates that if WMB is assumed to be the cause, then \rocrit~$\lesssim$~\rothresh, where \rothresh again is assumed to be a detection threshold.}

@@

We note that the \rocrit values we found in \S\ref{subsec:rossby} are smaller than the values of $\sim$2 found by \citet{vanSaders2016,vanSaders2019}, however, these results are sensitive to different \teff scales and different prescriptions for \taucz. We explored the \taucz formulae of \citet{BarnesKim2010} and \citet{Landin2010}, but these produce larger \taucz at fixed \teff than the \citet{CranmerSaar2011} relation, which exacerbates the \rocrit mismatch mentioned above. \edit1{As shown in \citet{See2021}, the \taucz relation of \citet{Amard2019} also lies above the \citet{CranmerSaar2011} relation, so adopting the newer \citet{Amard2019} relation would also not alleviate the \rocrit discrepancy.}

>>>>

To further explore whether our \rocrit inference procedure is biased by systematics in the data, we performed non-linear least-squares fits of two models to the CKS long-period pile-up and the \lamostmcq 90th percentile curve (both in the \teff range of 5800--6250~K) using the \texttt{curve\_fit} function in \texttt{scipy.optimize}. The first model assumes \rocrit = 2 and has one free parameter, a temperature offset added to the data. The second model assumes no temperature offset and allows \rocrit to vary (\rocrit being the only free parameter). The results of these fits are shown in Figure~\ref{fig:rocrit}. We found that for the CKS sample, particularly when using the more precise temperatures from \citet{Fulton2018}, that the $\Delta \chi^2$ between the \rocrit = 2 and variable \rocrit models was negligible. In other words, the \rocrit = 2 model fits the data well with a \teff shift of $\approx$~80--150~K depending on the \teff source used. For the \lamostmcq sample, there is weak support for the variable \rocrit model. However, as shown in Appendix~\ref{app:teff}, there is a strong systematic trend in the LAMOST temperatures when compared to other spectroscopic catalogs (particularly SPOCS). Correcting for these systematic trends weakens the support for the variable \rocrit model, and we speculate that temperature systematics bias the \rocrit inference procedure. Lending further support to this hypothesis is the observation of visibly different slopes for the long-period pile-up in the CKS vs. \lamostmcq samples (see Fig.~\ref{fig:percentiles}).\footnote{\edit1{While there is reasonably good agreement between the CKS long-period pile-up and the \lamostmcq 90th percentile curve, we observe in Figure~\ref{fig:percentiles} a systematic trend in which most CKS pile-up stars lie above the \lamostmcq curve for \teff~$\lesssim$~6000~K, and below it for \teff~$\gtrsim$~6000~K. This is consistent with there being a \teff-dependent systematic trend between the \teff scales of CKS and LAMOST, as discussed in Appendix~\ref{app:teff}.}}

@@

\subsection{Detection biases}
\label{subsec:detectionbias}
\edit1{To this point, we have not addressed the detection biases inherent to Kepler rotation period catalogs. As rotation periods become longer, they become more difficult to measure directly from rotational brightness modulations in Kepler time series. This is true for at least three reasons: (1) the Kepler baseline is finite, (2) photometric variation amplitude declines with increasing rotation period, and (3) photometric variations tend to become less periodic with increasing rotation period. This is important because, if a period detection threshold is Ro-dependent and the detection threshold Ro$_\mathrm{thresh}$ is close to the value \rocrit at which WMB becomes important, then the location of the pile-up in the catalogs considered here may depend sensitively on how detectability depends on Ro. For example, if Ro$_\mathrm{thresh}$ is only slightly larger than \rocrit and the long-period pile-up has some non-negligible width in period space (due either to astrophysics or measurement error), then it's possible the current data set only reveals the lower edge of the long-period pile-up.}

@@

\edit1{Finally, \citet{McQuillan2014} published weight parameters for each star, a metric which was designed to serve as a proxy for reliability, based on a star's autocorrelation function (ACF) properties and its position in \teff–\prot–LPH space, where LPH designates the local peak height of the ACF. Those authors selected $w>0.25$ for an acceptable compromise between real detections and false positives. In Figure~\ref{fig:weights}, we show that the long-period pile-up is apparent even for $w>0.3$, and we conclude that the majority of stars on the long-period pile-up are likely to be genuine detections and not an artifact of some detection threshold.}

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{./figures/weights.pdf}
    \caption{The \teff–\prot distribution of the \lamostmcq sample for different weight thresholds (indicated above each panel) computed by \citet{McQuillan2014}.}
    \label{fig:weights}    
\end{figure*}

@@

As discussed in \S\ref{subsec:rossby} and \S\ref{subsec:models}, the existence of a pile-up of stars with a constant Rossby number is consistent with expectations from the WMB model of \citet{vanSaders2016, vanSaders2019}. The location of the long-period pile-up coincides closely with the cluster of main-sequence stars with asteroseismic rotation rates determined by \citet{Hall2021}, \edit1{as well as the peak of the period distribution inferred by \citet{Masuda2021} from combining \vsini and stellar radii. We suggest that these features are one and the same.} Offsets in both observational and theoretical \teff scales between different studies \edit1{may help} explain the modest discrepancies between the asteroseismic sample and the samples considered here, as well as \edit1{differences} in the \rocrit values inferred here and those proposed in \citet{vanSaders2019}.

Furthermore, we note that there are important differences between the samples considered here and the \citet{Hall2021} asteroseismic sample. The present samples have rotation periods detected from photometric variations and, for a survey of finite baseline and sensitivity like Kepler, photometric rotation periods are harder to detect for slower rotators, smaller amplitude variations, and more stochastic variability patterns. Exacerbating the problem of detection, slower rotation tends to be accompanied by smaller-scale and less periodic variability. The asteroseismic sample, by comparison, is not biased against quiet, unspotted stars and is more likely to contain pile-up stars that our sample may have missed. \edit1{The same argument is true for the \citet{Masuda2021} sample.}

It is evident from Figure~\ref{fig:gap} that the long-period pile-up also coincides with a steep gradient in variability amplitude, such that a hidden population of pile-up stars may lie just beyond the edge of detectability. This effect, in addition to the \teff offsets mentioned above, can further bias our inference of \rocrit to lower values. Regardless of the biases mentioned above and the true value of \rocrit, the sample considered here clearly indicates that \edit1{a pile-up exists at the long-period edge} for stars with \teff~$\gtrsim$~\teffmin. \edit1{By construction, this pile-up occurs at a Rossby number lower than that associated with a detection edge.} \edit1{In other words, the fact that we observe a pile-up indicates that if WMB is assumed to be the cause, then \rocrit~$\lesssim$~\rothresh, where \rothresh again is assumed to be a detection threshold.}

@@

We note that the \rocrit values we found in \S\ref{subsec:rossby} are smaller than the values of $\sim$2 found by \citet{vanSaders2016,vanSaders2019}, however, these results are sensitive to different \teff scales and different prescriptions for \taucz. We explored the \taucz formulae of \citet{BarnesKim2010}, \citet{Landin2010}, \edit1{and \citet{Amard2019}}, but these produce larger \taucz at fixed \teff than the \citet{CranmerSaar2011} relation, which exacerbates the \rocrit mismatch mentioned above. \edit1{We also explored using a polynomial fit to the \taucz computed in \citet{vanSaders2019} for solar-metallicity stars, which produces a smaller \taucz at fixed \teff than the \citet{CranmerSaar2011} relation. However, even adopting this different \taucz prescription changes the inferred \rocrit upwards by only 15\%, not enough to obtain \rocrit~=~2.}
